---
title: "Findings: Disappearing Daycare"
execute:
  output: false
  echo: false
format:
  html:
    code-tools: true
    theme: cosmo
    toc: true
---

```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(tidycensus)
library(here)
library(sf)
library(lubridate)
library(tmap)
library(DT)
# Cache the zip code tabulation data we want from the Census 
options(tigris_use_cache = TRUE)
# Sets tmap to interactive mode
tmap_mode("view")
```

# Finding #1: Map 
- Missouri's childcare deserts in 2023, by Zip Code Tabulation Area
```{r}
# LOAD DATA

# Load clean and crosswalked data for all zctas from ChildCareAware survey
childcare_aware <- read_csv(here("data", "processed", "ChildCareAware_cleaned_and_crosswalked.csv")) %>% 
  mutate(zcta = as.character(zcta)) 

# Load clean and crosswalked data for all zctas from ChildCareAware survey, with all other zctas and populations baked in 
childcare_supply_and_demand_2019_2023 <-  read_csv(here("data", "processed", "childcare_supply_and_demand_2019_2023.csv")) %>% 
  mutate(zcta = as.character(zcta))

# Load all zctas not included in ChildCareAware survey 
# There are 1023 unique zctas in MO according to these crosswalk
mo_zctas <- read_excel(here("data", "raw", "ZIPCodetoZCTACrosswalk2021UDS.xlsx"), sheet = 1) %>% 
  clean_names() %>% 
  filter(state == "MO")

# 5 year ACS estimates pf population five and under for all zctas in the country
acs <- read_csv(here("data", "processed", "pop_zcta_5acs.csv"))

crrsa_startup_expansion <-  read_excel(here("data","raw", "CRRSA Startup and Expansion with Paid.xlsx")) %>% 
  clean_names()

# Load all relief data with cleaned zips and years for when and where money was dispersed to 
relief_funding <- read_csv(here("data", "processed", "all_relief_cleaned.csv")) %>% 
  mutate(zip = as.character(zip))


# Join crosswalk with relief funding zips to that each zip matches a zcta to correspond to desert data
relief_zcta <- relief_funding %>% 
  inner_join(mo_zctas, by = c("zip" = "zip_code"))
  


```
  
```{r}
supply_and_demand_2023 <- childcare_supply_and_demand_2019_2023 %>% 
  filter(year == "2023") %>% 
  mutate(ratio = pop_5_and_under / licensed_capacity) %>% 
  mutate(status = case_when(ratio < 3 ~ "non-desert", 
                            ratio >= 3 & ratio < 10 ~ "desert", 
                            ratio >= 10 ~ "extreme", 
                            is.na(ratio) & pop_5_and_under >= 50 ~ "desert",
                            is.na(ratio) & pop_5_and_under < 50 ~ "non-desert",
                            TRUE ~ "non-desert"))
  

map_df <- supply_and_demand_2023 %>% 
  mutate(status_num = case_when(status == "non-desert" ~ 0,
                                status == "desert" ~ 1,
                                status == "extreme" ~ 3)) %>% 
    select(zcta, status, status_num, ratio)

datawrapper_df <- supply_and_demand_2023 %>% 
  mutate(status_num = case_when(status == "non-desert" ~ 0,
                                status == "desert" ~ 1,
                                status == "extreme" ~ 3)) %>% 
  mutate(ratio = case_when(ratio >= 3 ~ round(ratio, digits = 0), 
                           TRUE ~ round(ratio, digits = 1))) %>% 
  mutate(dw = case_when(pop_5_and_under == 0 & childcare_programs >=1 ~ paste("0 children and", licensed_capacity, "slots"), 
                        is.na(pop_5_and_under) | ratio == 0 ~ paste("0 children and no childcare programs"), 
                        is.na(ratio) ~ paste(pop_5_and_under, "children and no childcare programs"), 
                        TRUE ~ paste("Child to slot ratio is", ratio, ":1")))

#write.csv(datawrapper_df, "datawrapper_map.csv")

```

```{r}
# JOIN CROSSWAKLED DATA WITH GEOM FOR MAPPING 
# Now we can use zctas to get geoms for mapping 

# Get the 2017-2021 5-year ACS ZTCAs for the whole US so that we can join these with zip codes and get geoms for mapping
# B09001_005 is a dummy variable, we just need the geoms from the call
zcta_geoms <- get_acs(geography = "zip code tabulation area", variables = "B09001_005", geometry = TRUE)  %>% 
  select(GEOID, geometry)

# Join the survey and non-survey data with geoms from Census for mapping 
# This join to zcta geoms from ACS data drops 5 values, but anti join shows they are all non-deserts and "point" zctas for PO boxes 
map_df_sf <- map_df %>% 
  inner_join(zcta_geoms, by = c("zcta" = "GEOID")) %>% 
  st_as_sf(sf_column_name = "geometry")

```

```{r}
#| output: true
# MAP DATA 
tm_shape(map_df_sf) +
  tm_polygons("status_num",
              palette = "Reds",
              legend.show = FALSE)
```

# Finding #2: Number of children in childcare deserts

- 202,318 children of five and under across Missouri live in zip codes areas that are childcare deserts. This is almost half of the 442,254 children five and under

```{r}
deserts_2023 <- supply_and_demand_2023 %>% 
  filter(status %in% c("desert", "extreme")) %>% 
  select(zcta, pop_5_and_under)

# 202,318 in 227 zctas that are childcare deserts
sum(deserts_2023$pop_5_and_under, na.rm = TRUE)


# Same five ZCTAs didn't join, but again, these are non-deserts PO boxes anyway
mo_zctas_with_pop <- mo_zctas %>% 
  left_join(acs, by = c("zcta" = "GEOID")) %>% 
  select(zcta, pop_5_and_under)

#  442,254 children under 5 or under across the state
sum(mo_zctas_with_pop$pop_5_and_under, na.rm=TRUE)

percent_in_deserts <- 202318/442254

```
### Population in childcare deserts 
```{r}
#| output: true
#| column: page
datatable(deserts_2023, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```
### Total population
```{r}
#| output: true
#| column: page
datatable(mo_zctas_with_pop, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```

# Finding #3: Pandemic changes in childcare 

- From 2019 to September 2020, the number of childcare programs decreased by 30%, leaving 80,000 more children in childcare deserts than the year before. During 2020, the number of children in childcare deserts was the highest it has reached in the last several years
- The number of programs bounced back sharply by six months later though, and since 2021, the number of children in childcare deserts has steadily towards where it was pre-pandemic

```{r}
yearly_deserts <- childcare_supply_and_demand_2019_2023 %>% 
  mutate(ratio = pop_5_and_under / licensed_capacity) %>% 
  mutate(status = case_when(ratio < 3 ~ "non-desert", 
                            ratio >= 3 ~ "desert", 
                            is.na(ratio) & pop_5_and_under >= 50 ~ "desert",
                            is.na(ratio) & pop_5_and_under < 50 ~ "non-desert",
                            TRUE ~ "non-desert")) %>% 
  filter(status == "desert") %>% 
  group_by(year, month) %>% 
  summarize(children_in_deserts = sum(pop_5_and_under))

num_programs_yearly <- childcare_aware %>% 
  group_by(year, month) %>% 
  summarize(num_program = sum(childcare_programs))

 
```
## Number of children in childcare deserts by month and year

```{r}
#| output: true
#| column: page
datatable(yearly_deserts, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```
## Number of childcare programs by month and year
```{r}
#| output: true
#| column: page
datatable(num_programs_yearly, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```

# Finding #4: Age breakouts of for startup, expansion, or enhancement

- Of programs that received CRRSA funding for startup, expansion, or enhancement, less than half as many slots were added for infant and toddlers as for preschool aged children

```{r}
age_breakouts <- crrsa_startup_expansion %>% 
  select(number_of_infant_toddler_slots_adding, number_of_preschool_slots_adding, number_of_school_age_slots_adding) %>% 
  mutate_all(~replace(., is.na(.), 0)) %>% 
  pivot_longer(number_of_infant_toddler_slots_adding:number_of_school_age_slots_adding, names_to = "age_group", values_to = "number")%>% 
  group_by(age_group) %>% 
  summarize(number = sum(number))


```

```{r}
#| output: true
#| column: page
datatable(age_breakouts, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```

# Finding #5: Funding data we recieved shows less money went to deserts


- Of the total of 114,681,667 dollars in the data we received, less than a third - 25,757,599 - of it went to childcare deserts
- Our 114,681,667 dollars is about half of the over 230,000,00 DESE has to spend in total through the several different funding sources
- 203 of 230 million has been awarded so far in total
- It looks like 170 million in total has been awarded from the grants we received, which more recent data would likely show. Our few months lag is the reason we don't have all of that 


```{r}
# Filter the capacity deserts to data from 2021 and 2022 and remove non-desert zctas, and aggregate to yearly
supply_and_demand_2021_2022 <- childcare_supply_and_demand_2019_2023 %>% 
  filter(year %in% c("2021", "2022")) %>% 
  group_by(zcta, year) %>% 
  summarize(pop_5_and_under = sum(pop_5_and_under), licensed_capacity = sum(licensed_capacity)) %>% 
  mutate(ratio = pop_5_and_under/licensed_capacity) %>% 
  mutate(status = case_when(ratio < 3 ~ "non-desert", 
                            ratio >= 3 ~ "desert", 
                            is.na(ratio) & pop_5_and_under >= 50 ~ "desert",
                            is.na(ratio) & pop_5_and_under < 50 ~ "non-desert",
                            TRUE ~ "non-desert")) 


# This join is for all funding together 
# Relief zcta has 380 unique ZCTAs, 661 unique year/zcta combos
relief_to_desert_join <- relief_zcta %>% 
  group_by(zcta, year) %>% 
  summarize(total = sum(amount)) %>% 
  left_join(supply_and_demand_2021_2022, by = c("zcta", "year")) 

```


```{r}
# The big reveal 
total_desert_relief <-  relief_to_desert_join %>% 
  group_by(status) %>% 
  summarize(total = sum(total))

portion_to_deserts <- 25757599/88924069
# Check total here with total from raw dataset to make sure now data went missing on joins 
# 114681667 in raw data matches exact amount here +1 for rounded 
check_total <- 25757599 + 88924069
```


```{r}
#| output: true
#| column: page
datatable(total_desert_relief, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```

```{r}
# This join is for funding my source 
relief_to_desert_type_join <- relief_zcta %>% 
  group_by(zcta, year, type) %>% 
  summarize(total = sum(amount)) %>% 
  left_join(supply_and_demand_2021_2022, by = c("zcta", "year"))  %>% 
  group_by(type, status) %>% 
  summarize(total = sum(total, na.rm = TRUE)) %>% 
  pivot_wider(names_from = status, values_from = total)

#write.csv(relief_to_desert_type_join, "datwrapper_stacked_bars.csv")

```

```{r}
#| output: true
#| column: page
datatable(relief_to_desert_type_join, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```
```{r}
dese_funds <- read_excel(here("data", "processed", "dese_funds_update.xlsx")) %>% 
  clean_names() %>% 
  mutate(awarded_approved_amount = as.numeric(if_else(awarded_approved_amount == "Pending", "0", awarded_approved_amount))) %>% 
  mutate(paid_amount_as_of_4_14_23 = as.numeric(paid_amount_as_of_4_14_23))


our_foia_data <- dese_funds %>% 
  filter(application_name %in% c("Enhancement", "Paycheck Protection", "Technical Business Assistance",
                                 "Expansion", "Startup", "Retention of Staff", "Annual Training Costs"))

foia_awarded <- sum(our_foia_data$awarded_approved_amount)

foia_paid <- sum(our_foia_data$paid_amount_as_of_4_14_23)

total_awarded <- sum(dese_funds$awarded_approved_amount)

total_paid <- sum(dese_funds$paid_amount_as_of_4_14_23)


our_portion <- 114681667/233314660

```
## All money in data sent us from DESE 
```{r}
#| output: true
#| column: page
datatable(dese_funds, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```
## Current numbers for our requests in data sent us from DESE 
```{r}
#| output: true
#| column: page
datatable(our_foia_data, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```
